LabMate: a prospectus for types for MATLAB

### 2024-04-11 writing/IMEKOWC2024/LabMate/paper.tex
[ writing ] adding

\newcommand{\append}{+\!\!\!+}
\newcommand{\hjux}[2]{{#1}|{#2}}
\newcommand{\vjux}[2]{\frac{#1}{#2}}

@@

Under the hood, LabMate translates selected well behaved MATLAB expressions
to typed expressions in its own core theory of types and values.
%
As matrices feature heavily in MATLAB code, the types of matrices
play a central role in this theory.
%
In particular, we acknowledge that the rows and columns of a matrix might pertain to distinct individual entities, rather as the rows ans columns of a spreadsheet do.
%
Internally, a matrix type looks like
\[\Matrix{R}{C}{E}{\rs}{\cs}
\]
%
We thus \param{} matrix types by two arbitrary `header' types $R$, for the individuals the rows concern, and $C$ for the columns. We may then give two lists, $\rs = [r_1,\ldots,r_m]\in\List{R}$ (so that each $r_i\in R$) and $\cs = [c_1,\ldots,c_n]\in\List{C}$ of those individuals. A \param{}d type $E(r,c)$ gives the type of entry for any combination of row and column individual, with $E(r_i,c_j)$ being the type of the entry $e_{i,j}$. In the diagram below, the $e$s inside the rectangle are the matrix entries computed by the MATLAB code. The $r$s and $c$s outside the rectangle are \remph{meta}data coming from the matrix type, known only to LabMate, and not available as data to MATLAB.
\[\begin{array}{r|ccccc|}
\; & c_1      & \cdots & c_j     & \cdots & c_n     \\
\hline
r_1    & e_{1,1}  &        & e_{1,j} &        & e_{1,n} \\
\vdots &          &        &         &        &         \\
r_i    & e_{i,1}  &        & e_{i,j} &        & e_{i,n} \\
\vdots &          &        &         &        &         \\
r_m    & e_{m,1}  &        & e_{m,j} &        & e_{m,n}  \\
\hline
\end{array}\]
To check that $\Matrix{R}{C}{E}{\rs}{\cs}$ is a valid type, LabMate must:
\begin{enumerate}
\item check that $R$ and $C$ are both valid types
\item check that $E(r,c)$ is a valid type whenever $r\in R$ and $c\in C$
\item check that $\rs\in\List{R}$ and $\cs\in\List{C}$
\end{enumerate}

@@

Firstly, a singleton entry matrix must have singleton header lists $[r]$ and $[c]$.
\[\begin{array}[b]{r|c|l}
\; & c \\
\cline{1-2}
r  & e\in E(r,c) & \quad \in \Matrix{R}{C}{E}{[r]}{[c]} \\
\cline{1-2}
\end{array} 
\]

@@

@@

Terms of matrix type in the core type theory are built up from
$1 \times 1$ matrices and pasting matrices horizontally and vertically.
%
Singleton matrices have the following typing rule
\[
  \infer{[e] : \Matrix{R}{C}{E}{[r]}{[c]}}
    {r : R
      &\quad
      c : C
      &\quad
      e : E(r, c)
    }
\]
%
while horizontal and vertical pasting horizontally have the following
typing rules:
\[
  \infer{\hjux{A}{B} : \Matrix{R}{C}{E}{\rs}{(cs_A \append cs_B)}}
  {
    A : \Matrix{R}{C}{E}{\rs}{cs_A}
    &\quad
    B : \Matrix{R}{C}{E}{\rs}{cs_B}
  }
\]
\[
  \infer{\vjux{A}{B} : \Matrix{R}{C}{E}{(\rs_A \append \rs_B)}{cs}}
  {
    A : \Matrix{R}{C}{E}{\rs_A}{cs}
    &\quad
    B : \Matrix{R}{C}{E}{\rs_B}{cs}
  }
\]
When comparing matrices for equality, we reassociate horizontal and
vertical pastings, so that $\hjux{\vjux{A}{C}}{\vjux{B}{D}}$ and
$\vjux{\hjux{A}{B}}{\hjux{C}{D}}$ indeed are considered equal.

The conditions for adding two matrices are given in Figure~\ref{fig:matrix_add} --- the matrices $A$ and $B$ can be added if all entry types $E(x, y)$ are
numeric (for example integers, or floating point numbers), and if $A$
and $B$ have the same matrix type.
%
%\[
%  \infer{A + B : \Matrix{R}{C}{E}{rs}{cs}}
%  {
%    x : R, y : C \vdash \isadd{E(x, y)}
%    &\quad
%    A : \Matrix{R}{C}{E}{rs}{cs}
%    &\quad
%    B : \Matrix{R}{C}{E}{rs}{cs}
%  }
%\]
%

>>>>

\newcommand{\append}{\mathop{+\!\!+}}
\newcommand{\hjux}[2]{{#1}|{#2}}
\newcommand{\vjux}[2]{\frac{#1}{#2}}

@@

Under the hood, LabMate translates selected well behaved MATLAB expressions
to typed expressions in its own core theory of types and values.
%
As matrices feature heavily in MATLAB code, the types of matrices
play a central role in this theory.
%
In particular, we acknowledge that the rows and columns of a matrix might pertain to distinct individual entities, rather as the rows ans columns of a spreadsheet do.
%
Internally, a matrix type looks like
\[\Matrix{R}{C}{E}{\rs}{\cs}
\]
%
We thus \param{} matrix types by two arbitrary `header' types $R$, for the individuals the rows concern, and $C$ for the columns. We may then give two lists, $\rs = [r_1,\ldots,r_m]\in\List{R}$ (so that each $r_i\in R$) and $\cs = [c_1,\ldots,c_n]\in\List{C}$ of those individuals. A \param{}d type $E(r,c)$ gives the type of entry for any combination of row and column individual, with $E(r_i,c_j)$ being the type of the entry $e_{i,j}$. In the diagram below, the $e$s inside the rectangle are the matrix entries computed by the MATLAB code. The $r$s and $c$s outside the rectangle are \remph{meta}data coming from the matrix type, known only to LabMate, and not available as data to MATLAB.
\[\begin{array}{r|ccccc|}
\; & c_1      & \cdots & c_j     & \cdots & c_n     \\
\hline
r_1    & e_{1,1}  &        & e_{1,j} &        & e_{1,n} \\
\vdots &          &        &         &        &         \\
r_i    & e_{i,1}  &        & e_{i,j} &        & e_{i,n} \\
\vdots &          &        &         &        &         \\
r_m    & e_{m,1}  &        & e_{m,j} &        & e_{m,n}  \\
\hline
\end{array}\]
To check that $\Matrix{R}{C}{E}{\rs}{\cs}$ is a valid type, LabMate must:
\begin{enumerate}
\item check that $R$ and $C$ are both valid types;
\item check that $E(r,c)$ is a valid type whenever $r\in R$ and $c\in C$;
\item check that $\rs\in\List{R}$ and $\cs\in\List{C}$.
\end{enumerate}

@@

Firstly, a singleton entry matrix must have singleton header lists $[r]$ and $[c]$.
\[\begin{array}[b]{r|@{}c@{}|l}
\; & c \\
\cline{1-2}
r  & e\in E(r,c) & \quad \in \Matrix{R}{C}{E}{[r]}{[c]} \\
\cline{1-2}
\end{array} 
\]

@@

Now that we have matrices, we must explain how to typecheck matrix operations.

To check if we may add matrices $A$ and $B$, LabMate must:
\begin{enumerate}
\item check that $A$ and $B$ both have the \remph{same} matrix type $\Matrix{R}{C}{E}{\rs}{\cs}$;
\item check that \remph{every} entry type $E(r,c)$ admits addition.
\end{enumerate}
In the short term, LabMate can get by with a hard-coded table of which types admit addition, but more flexible and general approaches to operator overloading (e.g., Haskell's `type classes'~\cite{wadler.blott}) are readily adoptable in due course.

@@

When comparing matrices for equality, we reassociate horizontal and
vertical pastings, so that $\hjux{\vjux{A}{C}}{\vjux{B}{D}}$ and
$\vjux{\hjux{A}{B}}{\hjux{C}{D}}$ indeed are considered equal.

%
%\[
%  \infer{A + B : \Matrix{R}{C}{E}{rs}{cs}}
%  {
%    x : R, y : C \vdash \isadd{E(x, y)}
%    &\quad
%    A : \Matrix{R}{C}{E}{rs}{cs}
%    &\quad
%    B : \Matrix{R}{C}{E}{rs}{cs}
%  }
%\]
%
